---
title: "What's in a name? The Gaelic of the Munros"
author: Emily Nordmann
date: '2026-06-25'
slug: gaelic-hill-names
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-location: left
execute:
  warning: false
  message: false
---

*This blog was written using Claude Opus 4.8*

A while back I [fell down a rabbit hole](../2025-08-31-munro-tidy-tuesday/) combining the Database of British and Irish Hills with route information scraped from [walkhighlands](https://www.walkhighlands.co.uk/). I've seen quite a few analyses of the most common Munro names and their meaning in Gaelic but I realised that the dataset from walkhighlands should allow me to go beyond what I've seen before.

I'm reusing the combined dataset from the previous post so please see that for details on how it was done. Whilst my wife and her family are Gaelic speakers, I am not and this blog uses Mark Jackson's [glossary for the Cambridge University Hillwalking Club](https://cuhwc.org.uk/resources/meanings-of-gaelic-words-commonly-seen-in-hill-names/) and Peter Drummond's *Scottish Hill Names* for the translations. 

```{r setup}
library(tidyverse)
library(stringi)
library(flextable)
library(ggthemes)
library(ggridges)
library(scales)
library(tidytext)
library(sf)
library(plotly)
library(rnaturalearth)

munro_dat <- read_csv("munros_combined.csv")

# The nature palette from the previous post
nature_6 <- c("#355E3B", "#4B4F58", "#4682B4", "#8E6C88", "#E07B39", "#BDB76B")
nature_5 <- c("#355E3B", "#4B4F58", "#4682B4", "#E07B39", "#BDB76B")
nature_4 <- c("#355E3B", "#4B4F58", "#4682B4", "#E07B39")
nature_2 <- c("#355E3B", "#4682B4")
nature_13 <- c("#355E3B", "#6B8E23", "#BDB76B", "#8B5A2B", "#D2B48C",
               "#87CEEB", "#4682B4", "#191970", "#7D7D7D", "#A9A9A9",
               "#8E6C88", "#E07B39", "#FFD700")
```

```{r theme-munro}
#| code-summary: "Show theme"
theme_munro <- function(base_size = 15) {
  theme_minimal(base_size = base_size) +
    theme(
      text             = element_text(colour = "#2B2B2B"),
      plot.title       = element_text(face = "bold", size = rel(1.45),
                                      lineheight = 1.05, margin = margin(b = 4)),
      plot.title.position = "plot",
      plot.subtitle    = element_text(size = rel(0.95), colour = "#5C5C5C",
                                      lineheight = 1.1, margin = margin(b = 16)),
      plot.caption     = element_text(size = rel(0.7), colour = "#8A8A8A",
                                      hjust = 0, margin = margin(t = 18)),
      plot.caption.position = "plot",
      axis.title.x     = element_text(size = rel(0.8), colour = "#5C5C5C",
                                      margin = margin(t = 8)),
      axis.title.y     = element_blank(),
      axis.text        = element_text(colour = "#2B2B2B"),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.major.x = element_line(colour = "#EDEDED", linewidth = 0.4),
      legend.position  = "top",
      legend.justification = "left",
      legend.title     = element_blank(),
      legend.text      = element_text(size = rel(0.9)),
      plot.margin      = margin(22, 26, 18, 22),
      plot.background  = element_rect(fill = "#FCFCFA", colour = NA),
      panel.background = element_rect(fill = "#FCFCFA", colour = NA)
    )
}

munro_cap <- "Source: walkhighlands · Database of British & Irish Hills v18.2"

# Representative hex for each Gaelic colour word, used in the colour plots
colour_hex <- c(
  "dearg / ruadh" = "#A33B2A", "glas" = "#7E8B6B", "liath" = "#9AA0A6",
  "geal" = "#E8E4D8", "buidhe" = "#D9A441", "dubh" = "#2B2B2B",
  "bàn" = "#D8D2C4", "gorm" = "#3E6E8E"
)
```

## The language of hills

Most Munro names follow the same grammar: a generic word telling you *what kind of hill* it is, usually followed (or preceded) by a qualifier describing it - a colour, a size, an animal, a saint. Here are the most common generic names.

```{r glossary}
tibble::tribble(
  ~Element, ~Meaning,
  "Beinn (Ben)", "The generic word for a mountain, especially a big one",
  "Sgùrr (Sgòrr, Sgòr)", "A jagged, pointed, rocky peak - mostly in the west",
  "Càrn (Cairn)", "A heap of stones; by extension a stony hill",
  "Stob", "A point or peak (literally a stake or stump)",
  "Meall", "A bare, rounded, lumpy hill",
  "Mullach", "A summit or top",
  "Aonach", "A high ridge",
  "Creag", "A crag or rockface",
  "Bidean (Spidean)", "A sharp pinnacle or peak",
  "Stùc", "A steep little peak or pinnacle",
  "Binnein", "An especially conical peak",
  "Maol", "A bare, bald hill"
) |>
  flextable() |>
  autofit()
```

The next step in the coding was to tag each Munro with its generic. This is fiddlier than it sounds because the word can be the first in the name (*Beinn Dearg*) or later when an adjective comes first (*Glas Bheinn*, *Geal Chàrn*). It also gets lenited - a softening of the opening consonant that Gaelic triggers in certain grammatical positions (after the article, for feminine nouns, after some adjectives) and which is written by adding an *h* in behind that first letter - so *beinn* becomes *bheinn* and *càrn* becomes *chàrn*, which a blunt search for the plain word would miss. And a good number are either anglicised (*Lochnagar*, *Braeriach*, *Schiehallion*) or built from rarer, more poetic words - *An Socach* (the snout), *A' Mhaighdean* (the maiden), *Am Bodach* (the old man). The code strips any accents, then searches the whole name for each generic, and everything that doesn't contain a recognised hill-word is categorised as "Other".

```{r classify}
# Normalise to an accent-free, lower-case key with punctuation as spaces
munro_dat <- munro_dat |>
  mutate(key = munro |>
           stri_trans_general("Latin-ASCII") |>
           str_to_lower() |>
           str_replace_all("[-']", " ") |>
           str_squish())

# Word-boundary regex for each generic. Order matters only for "Stob Binnein",
# where Stob should win, so Stob is tested before Binnein.
munro_dat <- munro_dat |>
  mutate(element = case_when(
    str_detect(key, "\\b(sgurr|sgor|sgorr|sgur)\\b")        ~ "Sgùrr",
    str_detect(key, "\\b(bidean|bidein|spidean|spidein)\\b") ~ "Bidean",
    str_detect(key, "\\b(stuc|stuchd|stucd|stac)\\b")        ~ "Stùc",
    str_detect(key, "\\b(stob)\\b")                          ~ "Stob",
    str_detect(key, "\\b(binnein|binnean)\\b")               ~ "Binnein",
    str_detect(key, "\\b(beinn|bheinn|ben)\\b")              ~ "Beinn",
    str_detect(key, "\\b(carn|charn|cairn|cairngorm|cairnwell)\\b") ~ "Càrn",
    str_detect(key, "\\b(meall|mheall)\\b")                  ~ "Meall",
    str_detect(key, "\\b(mullach)\\b")                       ~ "Mullach",
    str_detect(key, "\\b(aonach)\\b")                        ~ "Aonach",
    str_detect(key, "\\b(creag|chreag|craig)\\b")            ~ "Creag",
    str_detect(key, "\\b(maol|maoile|mhaoile)\\b")           ~ "Maol",
    str_detect(key, "\\b(tom)\\b")                           ~ "Tom",
    str_detect(key, "\\b(sail|saileag|saile)\\b")            ~ "Sàil",
    TRUE                                                     ~ "Other"
  ))

# A "scary" flag: route description mentions scrambling and/or exposure
munro_dat <- munro_dat |>
  mutate(scary = scramble_exposed %in% c("Scramble", "Exposed", "Both"))
```

*Beinn* is the most common all-purpose word for a mountain, followed by *Sgùrr* and *Càrn*. The "Other" group is proportionally large because it's where all the anglicised names and more unusual ones live.

```{r element-freq-polished}
#| code-summary: "Show polished version"
#| fig-width: 7
#| fig-height: 7
munro_dat |>
  count(element, sort = TRUE) |>
  mutate(element = fct_reorder(element, n)) |>
  ggplot(aes(n, element, fill = element)) +
  geom_col(width = 0.78) +
  geom_text(aes(label = n), hjust = -0.3, size = 4.2,
            fontface = "bold", colour = "#3A3A3A") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = colorRampPalette(nature_6)(15)) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL,
       title = "Which hill-words do the Munros use?",
       subtitle = "Generic Gaelic element in each of the 282 Munro names",
       caption = munro_cap) +
  theme_munro() +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
```

## Does the name tell you what you're climbing?

Many other people have done that analysis though, what I wanted to do was use the walkhighlands categories I made to see if the names can predict features like height or whether there's scary scrambling (still scared of heights). 

First, height. As in my previous blog, this only keeps words that have at least five Munros.

```{r height-ridges-polished}
#| code-summary: "Show polished version"
#| fig-width: 7
#| fig-height: 7
element_n <- munro_dat |> count(element) |> filter(n >= 5, element != "Other")

munro_dat |>
  semi_join(element_n, by = "element") |>
  group_by(element) |>
  mutate(mean_h = mean(height_dobih, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(height_dobih, fct_reorder(element, mean_h), fill = element)) +
  geom_density_ridges(quantile_lines = TRUE, quantile_fun = mean,
                      vline_linetype = "dashed",
                      colour = "white", linewidth = 0.4, scale = 1.5, alpha = 0.95) +
  scale_fill_manual(values = colorRampPalette(nature_6)(nrow(element_n)), guide = "none") +
  scale_y_discrete(expand = c(0.02, 0)) +
  scale_x_continuous(expand = c(0.01, 0)) +
  labs(x = "Height (m)", y = NULL,
       title = "Does the hill-word predict the height?",
       subtitle = "Distribution of heights per hill-word; dashed line = mean",
       caption = munro_cap) +
  theme_munro() +
  theme(panel.grid.major.x = element_line(colour = "#EDEDED", linewidth = 0.4))
```

There's not much in it because everything clusters around the 1,000 m mark, because that's roughly what it takes to be a Munro in the first place - the *Aonach's* edge it on average, but the differences are small and the Aonach's in particular are bimodal. 

What is much more predictive is whether the route contains a reference to scrambling or exposure:

```{r scary-rates-polished}
#| code-summary: "Show code"
#| fig-width: 7
#| fig-height: 7
scary_dat <- munro_dat |>
  semi_join(element_n, by = "element") |>
  group_by(element) |>
  summarise(n = n(),
            scary_pct = mean(scary) * 100,
            .groups = "drop") |>
  mutate(family = case_when(
    element %in% c("Sgùrr", "Bidean", "Stùc", "Stob", "Binnein") ~ "Pointed / rocky peaks",
    element %in% c("Càrn", "Meall", "Maol", "Mullach") ~ "Rounded or stony hills",
    TRUE ~ "General / other"
  ))

ggplot(scary_dat, aes(scary_pct, fct_reorder(element, scary_pct), fill = family)) +
  geom_col(width = 0.78) +
  geom_text(aes(label = paste0(round(scary_pct), "%")), hjust = -0.25,
            size = 4, fontface = "bold", colour = "#3A3A3A") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.13))) +
  scale_fill_manual(values = c("Pointed / rocky peaks" = "#E07B39",
                               "Rounded or stony hills"  = "#355E3B",
                               "General / other"        = "#4B4F58")) +
  labs(x = NULL, y = NULL,
       title = "Does the name tell you if it's scary?",
       subtitle = "Share of each word's routes that mention scrambling or exposure",
       caption = munro_cap) +
  theme_munro() +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
```

The words for sharp, pointed peaks - *Bidean* (83%), *Stùc* (80%) and *Sgùrr* (57%) - sit right at the top, while the words for rounded or stony hills - *Càrn* (31%) and *Meall* (32%) - sit at the bottom. *Beinn*, the generic catch-all, lands pleasingly in the middle.

The same caveat from the last post still applies that my "scary" flag comes from text-mining one route description per Munro for words like *scramble* and *exposed*, which is a blunt instrument that can't tell "an exposed ridge" from "exposed to the weather". And some of these groups are small - *Bidean* and *Stùc* are six and five hills respectively but the broad pattern across the bigger groups is fairly consistent.


## A linguistic map of Scotland

If the words describe the hills, and the hills differ by region (see my previous post), then the words should cluster geographically. Here's every Munro mapped by its generic, with pointed/rocky peak words highlighted in orange, rounded or stony hill words in green, and everything else greyed out.

```{r map-prep}
pointed_words <- c("Sgùrr", "Bidean", "Stùc", "Stob", "Binnein")
rounded_words  <- c("Càrn", "Meall", "Maol", "Mullach")

munros_coords <- munro_dat |>
  select(munro, region, element, xcoord, ycoord, height_dobih) |>
  drop_na(xcoord, ycoord) |>
  st_as_sf(coords = c("xcoord", "ycoord"), crs = 27700) |>
  st_transform(crs = 4326) |>
  mutate(longitude = st_coordinates(geometry)[, 1],
         latitude  = st_coordinates(geometry)[, 2]) |>
  st_drop_geometry()

uk_map_df <- rnaturalearth::ne_countries(scale = "large",
                                         country = "United Kingdom",
                                         returnclass = "sf") |>
  st_transform(4326) |>
  st_coordinates() |>
  as.data.frame() |>
  mutate(group = paste(L3, L2, L1, sep = "_"))

# Collapse to the main elements so the legend is readable
main_elements <- munro_dat |> count(element) |>
  filter(n >= 5, element != "Other") |> pull(element)

map_dat <- munros_coords |>
  mutate(element_grp = if_else(element %in% main_elements, element, "Other"))
```

```{r element-map-polished}
#| code-summary: "Show plot code"
#| fig-width: 7
#| fig-height: 7
map_dat2 <- munros_coords |>
  mutate(grp = case_when(
    element %in% pointed_words ~ "Pointed / rocky peaks",
    element %in% rounded_words ~ "Rounded or stony hills",
    TRUE                       ~ "Other words"))

ggplot() +
  geom_polygon(data = uk_map_df, aes(X, Y, group = group),
               fill = "#EAEAE6", colour = "#D5D5D0", linewidth = 0.3) +
  geom_point(data = filter(map_dat2, grp == "Other words"),
             aes(longitude, latitude), colour = "#CFCFCF", size = 1.6) +
  geom_point(data = filter(map_dat2, grp != "Other words"),
             aes(longitude, latitude, colour = grp), size = 2.6) +
  coord_quickmap(xlim = c(-8, -1.5), ylim = c(56.5, 58.6)) +
  scale_colour_manual(values = c("Pointed / rocky peaks"  = "#E07B39",
                                 "Rounded or stony hills" = "#355E3B")) +
  labs(title = "Jagged west, hilly east",
       subtitle = "Pointed/rocky peak words (orange) vs rounded or stony hill words (green)",
       caption = munro_cap, colour = NULL) +
  theme_munro() +
  theme(axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())
```

The pattern becomes clearer when you break it down by region.

```{r region-shares}
munro_dat |>
  group_by(region) |>
  summarise(
    n                   = n(),
    `Pointed / rocky %` = mean(element %in% pointed_words) * 100,
    `Rounded or stony %` = mean(element %in% rounded_words) * 100,
    .groups = "drop"
  ) |>
  filter(n >= 8) |>
  arrange(desc(`Pointed / rocky %`)) |>
  flextable() |>
  colformat_double(j = c("Pointed / rocky %", "Rounded or stony %"), digits = 0) |>
  set_header_labels(n = "Munros") |>
  autofit()
```

The Isle of Skye is dominated by pointed/rocky peak words and has almost no rounded or stony hill names - the Cuillin are jagged rock and the names reflect that. The Cairngorms are the mirror image, with the highest share of rounded or stony words of any region. The rough north-west - Kintail, Torridon, Ullapool - leans pointed; Perthshire has almost none.

```{r region-bars-polished}
#| code-summary: "Show polished version"
#| fig-width: 7
#| fig-height: 7
munro_dat |>
  mutate(family = case_when(
    element %in% pointed_words ~ "Pointed / rocky peaks",
    element %in% rounded_words ~ "Rounded or stony hills",
    TRUE                       ~ NA_character_
  )) |>
  filter(!is.na(family)) |>
  count(region, family) |>
  group_by(region) |>
  filter(sum(n) >= 3) |>
  ungroup() |>
  ggplot(aes(n, fct_reorder(region, n, sum), fill = family)) +
  geom_col(width = 0.74) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c("Pointed / rocky peaks"  = "#E07B39",
                               "Rounded or stony hills" = "#355E3B")) +
  labs(x = "Number of Munros", y = NULL,
       title = "Jagged west, hilly east",
       subtitle = "Pointed/rocky peak words vs rounded or stony hill words, by region",
       caption = munro_cap) +
  theme_munro()
```

## Munros and Corbetts

The Munros are hills over 3,000 ft whilst their smaller siblings are the Corbetts (2,500-3,000 ft). 

```{r munro-corbett-polished}
#| code-summary: "Show polished version"
#| fig-width: 7
#| fig-height: 7
corbett <- read_csv("corbett.csv")

# Reusable version of the classifier so it can run over either list
classify_element <- function(x) {
  key <- x |>
    stri_trans_general("Latin-ASCII") |>
    str_to_lower() |>
    str_replace_all("[-']", " ") |>
    str_squish()
  case_when(
    str_detect(key, "\\b(sgurr|sgor|sgorr|sgur)\\b")        ~ "Sgùrr",
    str_detect(key, "\\b(bidean|bidein|spidean|spidein)\\b") ~ "Bidean",
    str_detect(key, "\\b(stuc|stuchd|stucd|stac)\\b")        ~ "Stùc",
    str_detect(key, "\\b(stob)\\b")                          ~ "Stob",
    str_detect(key, "\\b(binnein|binnean)\\b")               ~ "Binnein",
    str_detect(key, "\\b(beinn|bheinn|ben)\\b")              ~ "Beinn",
    str_detect(key, "\\b(carn|charn|cairn|cairngorm|cairnwell)\\b") ~ "Càrn",
    str_detect(key, "\\b(meall|mheall)\\b")                  ~ "Meall",
    str_detect(key, "\\b(mullach)\\b")                       ~ "Mullach",
    str_detect(key, "\\b(aonach)\\b")                        ~ "Aonach",
    str_detect(key, "\\b(creag|chreag|craig)\\b")            ~ "Creag",
    str_detect(key, "\\b(maol|maoile|mhaoile)\\b")           ~ "Maol",
    str_detect(key, "\\b(tom)\\b")                           ~ "Tom",
    str_detect(key, "\\b(sail|saileag|saile)\\b")            ~ "Sàil",
    TRUE                                                     ~ "Other"
  )
}

compare <- bind_rows(
  tibble(name = munro_dat$munro, list = "Munro (n = 282)"),
  tibble(name = corbett$corbett, list = "Corbett (n = 222)")
) |>
  mutate(element = classify_element(name)) |>
  count(list, element) |>
  group_by(list) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()

keep <- c("Sgùrr", "Stob", "Bidean", "Stùc", "Beinn",
          "Càrn", "Meall", "Creag", "Aonach", "Mullach", "Other")

compare |>
  filter(element %in% keep) |>
  ggplot(aes(pct, fct_reorder(element, pct, .fun = max), fill = list)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.64) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_dodge(width = 0.72), hjust = -0.2,
            size = 3.3, colour = "#3A3A3A") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c("Munro (n = 282)"   = "#355E3B",
                               "Corbett (n = 222)" = "#C9A227")) +
  labs(x = NULL, y = NULL,
       title = "Munros vs Corbetts",
       subtitle = "Munros favour pointed-peak words; Corbetts are more likely to use Beinn",
       caption = munro_cap) +
  theme_munro() +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
```

The pointed, rocky words are much more Munro than Corbett - *Sgùrr* names 17% of Munros but only 10% of Corbetts, and across all the sharp-peak words together (*Sgùrr*, *Stob*, *Bidean*, *Stùc*) the Munros come to 28% against the Corbetts' 15%. The Corbetts, meanwhile, are much more likely to use *Beinn* - "a hill" - which accounts for 36% of them against 24% of Munros, and they have a bigger "Other" pile.

## Colour me Munro

The generic tells you what kind of hill it is whilst the qualifier tells you what it looks like. By far the most common qualifier is *mòr* (big), which appears on 26 Munros, with its opposite *beag* (small) trailing on four. But the qualifiers I find interesting are the colours.

```{r colour-freq-polished}
#| code-summary: "Show code"
#| fig-width: 7
#| fig-height: 7
colour_words <- tribble(
  ~colour, ~meaning, ~pattern,
  "dearg / ruadh", "red, rust",        "\\b(dearg|dhearg|deirg|ruadh|ruaidh)\\b",
  "glas",          "grey-green",       "\\b(glas|ghlas|ghlais)\\b",
  "liath",         "grey",             "\\b(liath|leith)\\b",
  "geal",          "white, bright",    "\\b(geal|gheal)\\b",
  "buidhe",        "yellow",           "\\b(buidhe|bhuidhe)\\b",
  "dubh",          "black",            "\\b(dubh|dhubh|duibh|dhuibh)\\b",
  "bàn",           "white, pale",      "\\b(ban|bhan|bhain)\\b",
  "gorm",          "blue",             "\\b(gorm|ghorm)\\b"
)

colour_counts <- colour_words |>
  mutate(n = map_int(pattern, \(p) sum(str_detect(munro_dat$key, p)))) |>
  mutate(label = paste0(colour, " (", meaning, ")"))

ggplot(colour_counts, aes(n, fct_reorder(label, n), fill = colour)) +
  geom_col(width = 0.76, colour = "grey75", linewidth = 0.3) +
  geom_text(aes(label = n), hjust = -0.4, size = 4,
            fontface = "bold", colour = "#3A3A3A") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = colour_hex, guide = "none") +
  labs(x = NULL, y = NULL,
       title = "The colours of the Munros",
       subtitle = "Munro names containing each Gaelic colour word",
       caption = munro_cap) +
  theme_munro() +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
```

Red wins - *dearg* and *ruadh* together name 15 Munros, more than any other colour. Then come the greys and grey-greens, *glas* and *liath*. What there is almost none of is *gorm*, blue. The explanation is that the Gaelic colour wheel doesn't map onto the English one: *glas* covers a  range from grey through to the green of grass and the grey-green of the sea, and *gorm* can shade into green too. The hills aren't painted in our colours, and translating them flattens something. A *Càrn Gorm* isn't quite a "blue cairn", and a *Beinn Ghlas* isn't quite a "grey mountain" because language is wondrous. 

### Red for danger

I then looked at whether certain colours were associated with scary features - the numbers for each group are so small that this isn't the most robust but there does seem to be some relationship with red. Red names tend to mark bare, scree-covered rock which lends itself to scrambling - the most notorious example being *Sgùrr Dearg*, the red peak that carries the Inaccessible Pinnacle. The grey-greens, by contrast, are the colours of grass and vegetated slope, which is to say the colours of a nice walk. 

```{r colour-scary-polished}
#| code-summary: "Show code"
#| fig-width: 7
#| fig-height: 7
colour_scary <- colour_words |>
  mutate(
    n = map_int(pattern, \(p) sum(str_detect(munro_dat$key, p))),
    scary_pct = map_dbl(pattern, \(p) mean(munro_dat$scary[str_detect(munro_dat$key, p)]) * 100),
    label = paste0(colour, " (n = ", n, ")")
  )

ggplot(colour_scary, aes(scary_pct, fct_reorder(label, scary_pct), fill = colour)) +
  geom_col(width = 0.76, colour = "grey75", linewidth = 0.3) +
  geom_vline(xintercept = mean(munro_dat$scary) * 100,
             linetype = "dashed", colour = "grey45") +
  geom_text(aes(label = paste0(round(scary_pct), "%")), hjust = -0.3,
            size = 3.8, fontface = "bold", colour = "#3A3A3A") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14)), limits = c(0, 100)) +
  scale_fill_manual(values = colour_hex, guide = "none") +
  labs(x = "Routes mentioning scrambling or exposure (%)", y = NULL,
       title = "Do the red hills bite?",
       subtitle = "By colour word. Dashed line = overall average (46%). Tiny samples",
       caption = munro_cap) +
  theme_munro()
```


