---
title: This is your thesis
author: ''
date: '2021-11-24'
slug: this-is-your-thesis
subtitle: 'For Kathleen'
lastmod: '2021-11-24T00:01:44Z'
featured: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

library(tidyverse)
library(tidytext)
library(readtext)
library(igraph)
library(ggraph)
library(wordcloud)
 
options(scipen = 999)

pride <- c("#613915", "#000000", "#FF0018", "#FFA52C", "#FFFF41", "#008018", "#0000F9", "#86007D")

trans <- c("#73D7EE","#73D7EE", "#FFAFC7", "#FFFFFF")


ch1 <- readtext("chapters/chapter_1.pdf")
ch2 <- readtext("chapters/chapter_2.pdf")
ch3 <- readtext("chapters/chapter_3.pdf")
ch4 <- readtext("chapters/chapter_4.pdf")
ch5 <- readtext("chapters/chapter_5.pdf")
ch6 <- readtext("chapters/chapter_6.pdf")
ch7 <- readtext("chapters/chapter_7.pdf")
ch8 <- readtext("chapters/chapter_8.pdf")

thesis_all <- bind_rows(ch1, ch2, ch3, ch4, ch5, ch6, ch7, ch8)

thesis <- thesis_all %>%
    unnest_tokens(output = word, input = text)

chapter_names <- as_labeller(c("chapter_1.pdf" = "Chapter 1",
                               "chapter_2.pdf" = "Chapter 2",
                               "chapter_3.pdf" = "Chapter 3",
                               "chapter_4.pdf" = "Chapter 4",
                               "chapter_5.pdf" = "Chapter 5",
                               "chapter_6.pdf" = "Chapter 6",
                               "chapter_7.pdf" = "Chapter 7",
                               "chapter_8.pdf" = "Chapter 8"))
```

```{r}
custom_stop <-tibble(word =  c("id", "1"))



tidy_thesis <- thesis %>%
  anti_join(stop_words) %>%
  anti_join(custom_stop)

full_count <- thesis %>% count()

unique_words <- thesis %>% 
  group_by(word) %>%
  count() %>%
  nrow()

words_per_chapter <- thesis %>%
  group_by(doc_id) %>%
  count() %>%
  rename(words = n)

unique_words <- thesis %>% 
  group_by(doc_id, word) %>%
  count() %>%
  ungroup(word) %>%
  count()%>%
  rename(unique = n) %>%
  inner_join(words_per_chapter) %>%
  mutate(prop_unique = (unique/words)*100) %>%
  ungroup()

unique_total <- unique_words %>%
  summarise(total = sum(unique))


```

Dear Kathleen,

This is your thesis. As interpreted by a nerd.

Your thesis has `r full_count$n[1]` words in total. Don't freak out if this is different to something you wrote on an official form, R will count every word in every table and footnote. What matters is that it is `r full_count$n[1] - 52000` words longer than your wife's thesis.

<br>

```{r}
thesis %>%
  group_by(doc_id) %>%
  count() %>%
  ggplot(aes(x = doc_id, y = n)) +
  geom_bar(stat = "identity", fill = pride, colour = "black") +
  theme_minimal() +
  labs(title = "Word count by chapter", x = "Chapter", y = "Words") +
  scale_x_discrete(labels = c("1", "2", "3", "4", "5", "6", "7", "8"))
```

<br>

In your thesis you used `r unique_total$total[1]` unique words and numbers. But which chapter had the highest proportion of unique words?

<br>

```{r}
unique_words %>%
  ggplot(aes(x = doc_id, y = prop_unique)) +
  geom_bar(stat = "identity", fill = pride, colour = "black") +
  theme_minimal() +
  labs(title = "Percentage of unique words by chapter", x = "Chapter", y = "Percent") +
  scale_x_discrete(labels = c("1", "2", "3", "4", "5", "6", "7", "8")) +
  scale_y_continuous(breaks = seq(from = 0, to = 30, by = 5))
```

```{r}
reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  new_x <- paste(x, within, sep = sep)
  stats::reorder(new_x, by, FUN = fun)
}

#' @rdname reorder_within
#' @export
scale_x_reordered <- function(..., sep = "___") {
  reg <- paste0(sep, ".+$")
  ggplot2::scale_x_discrete(labels = function(x) gsub(reg, "", x), ...)
}


#' @rdname reorder_within
#' @export
scale_y_reordered <- function(..., sep = "___") {
  reg <- paste0(sep, ".+$")
  ggplot2::scale_y_discrete(labels = function(x) gsub(reg, "", x), ...)
}
```

Chapter 1 would win at Scrabble.

<br>

But what are all those words? Here are the top words used per chapter, excluding "stop words" which are common function words like "and", "the", "it", etc.

<br>

```{r}
tidy_thesis %>%
  group_by(doc_id, word) %>%
  count() %>%
  arrange(-n) %>%
  ungroup(word) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5) %>%
  arrange(doc_id) %>%
  ggplot(aes(x = reorder_within(word, n, doc_id), y = n, fill = doc_id)) +
  geom_bar(stat = "identity", show.legend = F, colour = "black") +
  scale_x_reordered() +
  facet_wrap(~doc_id, scales = "free", nrow = 4, labeller = chapter_names) +
  coord_flip() +
  theme_minimal() +
  scale_fill_manual(values = pride) +
  labs(title = "Top five words used per chapter", x = NULL, y = NULL)

```

<br>

If you're so nerdily inclined, you can also look at bigrams, pairs of consecutive words (excluding stop words).

<br>

```{r}
thesis_bigrams <- thesis_all %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2)

bigrams_separated <- thesis_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)

# new bigram counts:
bigram_counts <- bigrams_filtered %>% 
  count(word1, word2, sort = TRUE) %>%
  ungroup()

bigrams_united <- bigrams_filtered %>%
  unite(bigram, word1, word2, sep = " ")

bigram_tf_idf <- bigrams_united %>%
  count(doc_id, bigram) %>%
  bind_tf_idf(bigram, doc_id, n) %>%
  arrange(desc(tf_idf))

bigrams_united %>%
  group_by(doc_id, bigram) %>%
  count() %>%
  arrange(-n) %>%
  ungroup(bigram) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5) %>%
  arrange(doc_id) %>%
  ggplot(aes(x = reorder_within(bigram, n, doc_id), y = n, fill = doc_id)) +
  geom_bar(stat = "identity", show.legend = F, colour = "black") +
  scale_x_reordered() +
  facet_wrap(~doc_id, scales = "free", nrow = 4, labeller = chapter_names) +
  coord_flip() +
  theme_minimal() +
  scale_fill_manual(values = pride) +
  labs(title = "Top five bigrams used per chapter", x = NULL, y = NULL)


```

<br>

Honestly, I'm not completely sure that this graph is at all useful, but it looks nice doesn't it?

<br>

```{r}
# filter for only relatively common combinations
bigram_graph <- bigram_counts %>%
  filter(n > 20) %>%
  graph_from_data_frame()

a <- grid::arrow(type = "closed", length = unit(.15, "inches"))

set.seed(12042018)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), show.legend = FALSE,
                 arrow = a, end_cap = circle(.07, 'inches')) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
  theme_void()
```

<br>

Who doesn't love a word cloud? Here's your top 50 words of the entire thesis (excluding stop words).

<br>

```{r}
tidy_thesis %>%
  anti_join(stop_words) %>%
  count(word) %>%
  with(wordcloud(word, n, max.words = 50))
```

<br>

What about the most positive and negative words?

<br>

```{r}
library(reshape2)

tidy_thesis %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  acast(word ~ sentiment, value.var = "n", fill = 0) %>%
  comparison.cloud(colors = c("gray20", "#F446FF"),
                   max.words = 50)
```

<br>

Finally, because I love you, a pie chart.

LOOK HOW HARD IT IS TO READ.

<br>

```{r}

ggplot(words_per_chapter, aes(x = "", 
                      y = words, 
                      fill = doc_id)) +
  geom_bar(width = 1, stat="identity") + 
  coord_polar("y", start = 0) +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank()
  ) +
  theme_void() +
  scale_fill_manual(name = "", labels = chapter_names, values = pride) +
  labs(title = "Word count by chapter")

```

<br>

Here's a clearer one.

<br>

```{r}
proud <- tibble(feeling = c("Pride", "Love", "Admiration"), 
                n = c(1,1,1))

ggplot(proud, aes(x = "", 
                      y = n, 
                      fill = feeling)) +
  geom_bar(width = 1, stat="identity") + 
  coord_polar("y", start = 0) +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank()
  ) +
  theme_void() + 
  labs(title = "What I feel when I look at you", fill = NULL) +
  scale_fill_viridis_d()

```
